---
title: "Metagenomic data analysis pipeline"
subtitle: "Voor het verwerken en analyseren van Oxford Nanopore Technologies’ MinION long-read, single-end shotgun sequencing data"
author: "Moreno Serafino"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


# Introduction

[blabla]

## Contents

[inhoudsopgave]

## Library setup:

```{r}

# load libraries:
# ...

```


## Environment setup:

```{bash, echo=TRUE, eval=FALSE}

# Update conda; echo "y" to automatically answer prompt:

echo "y" | conda update -n base conda

# activate conda meta environment:

conda activate meta

```


# Raw sequencing data overview

To create a summary report of the raw data for general purpose, run the FASTQ_summarizer.sh script. The "longest read length" result can provide insight in the type of sequencing data (long read vs. short read). Note that this RMarkdown file is for long read data analysis only.

cmd usage:  
bash FASTQ_summarizer.sh  
-I /home/data/projecticum/SW/raw_data  
-O ~/Metagen_SW/analyse  

```{bash, echo=TRUE, eval=FALSE}

#!/bin/bash


## step 0: set locale to avoid awk warning

export LC_ALL=C


## step 1: script description and usage:

print_usage() {
  printf "\nDescription of FASTQ_summarizer.sh: 
  
  This script is used for insightful summary of selected FASTQ files.
  
  Usage: 
  -I input directory with fastq.gz files
  -O output directory (created through script)
  -h for help

  Example cmd line:
	bash FASTQ_summarizer.sh -I path/to/FASTQ/data -O new/ouput/directory\n\n"
}

## step 2: flag definition:

while getopts I:O:h flag; do
  case "${flag}" in
    I) indir="${OPTARG}"  ;;
    O) outdir="${OPTARG}" ;;
    h) print_usage
       exit 1             ;;
    *) echo -e "\nPlease provide valid options (-I and -O are required)."
       echo -e "Use cmd \"bash FASTQ_summarizer.sh -h\" for instructions.\n"
       exit 1             ;;
  esac
done


## step 3.1: check whether flags have been provided by the user:

if [ -z "$indir" ] || [ -z "$outdir" ]; then
  echo -e "\nPlease provide valid options (-I and -O are required)."
  echo -e "Use cmd \"bash FASTQ_summarizer.sh -h\" for instructions.\n"
  exit 1
fi


## step 3.2: check whether directory -I exists and contains data:

if [ ! -d "$indir" ]; then
  echo -e "\nThe provided input directory does not exist."
  echo "Use cmd \"bash FASTQ_summarizer.sh -h\" for instructions.\n"
  exit 1
fi


## step 3.3: check whether directory -O exists, create if necessary:

if [ ! -d "$outdir" ]; then
  echo -e "\nCreating output directory: $outdir"
  mkdir -p "$outdir"
    exit 1
fi


## step 4: empty variables:

total_files=0
total_length=0
read_lengths=()
sorted_lengths=()


## step 5: analysis:

echo -e "\nAnalizing ..."

for file in "$indir"/*.fastq.gz; do
  current_read_length=$(awk 'NR%4==2 {sum+=length}END{print sum}' "$file")
  total_files=$((total_files + 1))
  total_length=$((total_length + current_read_length))
  read_lengths+=("$current_read_length")
done

shortest_read=$(
  { for file in "$indir"/*.fastq.gz; do
      gunzip -c "$file" | awk 'NR%4==2 {print length, $0}'
    done | sort -n | head -n1 | cut -d ' ' -f 2- | wc -c
  } 2>/dev/null
)

longest_read=$(
  { for file in "$indir"/*.fastq.gz; do
      gunzip -c "$file" | awk 'NR%4==2 {print length, $0}'
    done | sort -n | tail -n1 | cut -d ' ' -f 2- | wc -c
  } 2>/dev/null
)

## step 6: calculations:

echo -e "\nCalculating ..."

total_reads=0
for file in "$indir"/*fastq.gz; do
  lines=$(wc -l < "$file")
  total_reads=$((total_reads + lines /4))
done
average_length=$(($total_length / $total_reads))


## step 7: presentation:

outfile="SW_FASTQ_summary.txt"

{
echo -e "\nSummary analysis of FASTQ files found in $indir directory." 
echo "Script used: FASTQ_summarizer.sh."
echo "Input directory used: $indir"
echo "Script execution: $(date)"
echo -e "\n\n#===Report=======#\n\nTotal number of FASTQ files used as input:" 
echo "$total_files"
echo -e "\nTotal number of reads:"
echo "$total_reads"
echo -e "\nTotal combined read length:"
echo "$total_length"
echo -e "\nShortest read length:"
echo "$shortest_read"
echo -e "\nLongest read length:"
echo "$longest_read"
echo -e "\nAverage read length:"
echo "$average_length"
echo -e "\n#===Fin==========#"
} > "$outdir/$outfile"

echo -e "\nSummary report added to: $outdir/$outfile.\n"

```

```{bash, engine.opts='-i', echo=TRUE, eval=TRUE}

# Show summary:

cat ~/Metagen_SW/analyse/analyse_misc/SW_FASTQ_summary.txt

```

## FastQC report

The FastQC tool will be used to get a quick qualitative overview using the Phred-scores in the dataset. The following script will take the concatenated fastq.gz file, analyze the data and create an html report. The report consists of a number of graphs, the first and most important will be the per base quality score. These boxplots represent the quality values of the entire input per basecall 'location' in the reads, i.e. the quality scores of the first base of each read combined is represented by the first box. Results of this analysis can later be used to trim reads in order to obtain a subset with higher average quality.

```{bash, echo=TRUE, eval=FALSE}
#!/bin/bash

### pseudocode ###

## stap 1: beschrijf de functie en gebruikswijze van het script.
## stap 2: definieer de invoeropties.
## stap 3: voer de FastQC analyse uit d.m.v. fastqc command.


### code ###

## step 1: script description and usage:

print_usage() {
  printf "\nDescription of FastQC_reporter.sh: 
  
  This script can be used to create FastQC reports for fastq.gz files.
  
  Usage: 
  -I input directory with fastq.gz files
  -O output directory (created through script)
  -h for help

  Example cmd line:
	bash FASTQC_reporter.sh 
	-I path/to/FASTQ/data 
	-O new/ouput/directory (can be created through script)\n\n"
}


## step 2: flag definition:

while getopts I:O:h flag; do
  case "${flag}" in
    I) indir="${OPTARG}"  ;;
    O) outdir="${OPTARG}" ;;
    h) print_usage
       exit 1             ;;
    *) echo -e "\nPlease provide valid options (-I and -O are required)."
       echo -e "Use cmd \"bash FASTQC_reporter.sh -h\" for instructions.\n"
       exit 1             ;;
  esac
done


## step 3: FastQC analysis report:

for file in "$indir"/*.fastq.gz
do
    echo "Running FastQC on $file..."
    fastqc "$file" -o "$indir"
done

echo "FastQC analysis complete. Please check $outdir."

```



## FASTQ analysis with NanoPlot

To generate one NanoPlot summary for the entire FASTQ dataset, first concatenate all FASTQ files into one combined file.

```{bash, echo=TRUE, eval=FALSE}

cat /home/data/projecticum/SW/raw_data/*.fastq.gz > ~/Metagen_SW/raw_data/merged.fastq.gz

```


Then use the FASTQ_NanoPlotter.sh script to create a complete summary report.

cmd usage:  
bash FASTQ_NanoPlotter.sh  
-I ~/Metagen_SW/raw_data  
-O ~/Metagen_SW/analyse  


```{bash, echo=TRUE, eval=FALSE}

#!/bin/bash

# step 1: script description and usage:

print_usage () {
  printf "\nDescription of FASTQ_NanoPlotter.sh: 
  
  This script is used to create NanoPlots of FASTQ files. 
  These plots serve as quality control.
  Designed for long read data only!
  
  Usage: 
  -I input directory with fastq.gz files
  -O output directory (created through script)
  -h for help

  Example cmd line:
	bash FASTQ_NanoPlotter.sh -I path/to/FASTQ/data -O new/ouput/directory\n\n"
}


## step 2: flag definition:

while getopts I:O:h flag; do
  case "${flag}" in
    I) indir="${OPTARG}"  ;;
    O) outdir="${OPTARG}" ;;
    h) print_usage
       exit 1             ;;
    *) echo -e "\nPlease provide valid options (-I and -O are required)."
       echo -e "Use cmd \"bash FASTQ_NanoPlotter.sh -h\" for instructions.\n"
       exit 1             ;;
  esac
done


## step 3: check whether directory -O exists, create if necessary:

if [ ! -d "$outdir" ]; then
  echo -e "\nCreating output directory: $outdir\n"
  mkdir -p "$outdir"
fi


## step 4: NanoPlotting:

for FASTQ in "$indir"/*.fastq.gz; do
  file_name=$(basename "${FASTQ%.fastq.gz}")
  echo "Processing file: $FASTQ"
  NanoPlot --fastq_rich $FASTQ -o "$outdir/$file_name"
  
  #   xdg-open "$outdir/$file_name/NanoPlot-report.html" 
  #-- dit zou beter (algemener) zijn dan volgende line, maar werkt niet. 
  
  xdg-open https://daur.rstudio.hu.nl/s/c310e7760b90faf49cb24/files/Metagen_SW/analyse/merged_test_10/NanoPlot-report.html 
  #-- dit werkt wel.

## step 5: folder cleanup:

echo -e "\nRemoving unnecessary files ..."
find "$outdir" -type f ! -name "NanoPlot-report.html" -exec rm {} +
echo -e "\nNanoPlot created and folder cleaned up."
done

echo -e "\nOpening NanoPlot summary in browser.\n"

```

With this script, a NanoPlot summary document is created as an html file and opened automatically in the browser.

The NanoPlot function has calculated many aspects of the raw FASTQ data. Since the script specified the input as "fastq_rich", the report even generates plots concerning metadata such as the activity of each nanopore channel. The most important information in this document however is in the summary table:

- Mean read length should show that the input data is indeed produced from long read sequencing, since Nanopore is a long read NGS technique (as opposed to 100-150 bp illumina short read sequencing). This is important for downstream analysis.
- Mean read quality takes the average basecall Phred quality score over each read, and then gives the average score of all reads.
- \>Qx shows the proportion of reads with Phred quality score > x. In the SW_Metagen project case, >Q7 shows 100% of all reads, and >Q10 shows <100%. This suggests a preselection of raw data with a minimum Phred quality score cutoff somewhere between 7 and 10.

```{bash, echo=TRUE, eval=FALSE}

#-- HIER KOMT EEN SCRIPT. SCRIPT MAKEN VOOR KRAKEN2.
#-- command in de trend van: kraken2 --db /home/daur2/metagenomics/minikraken2_v2_8GB_201904_UPDATE/ --threads 2 --gzip-compressed --output ~/Metagen_SW/analyse/Kraken2_test_10/test_10.kraken --report ~/Metagen_SW/analyse/Kraken2_test_10/test_10.report --use-names ~/Metagen_SW/raw_data/merged_test_10.fastq.gz

```


